---
title: "PDDS_Query"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)
library(ggpubr)


#Finds PDDS values and median scores per year for all PROMOTE participants
  #Also finds EDSS values stored in the PROMOTEDatabase

#Files used: PROMOTEDatabase, PROMOTE Longitudinal, Legacy, Missing IDs, Legacy SNQ, sensor, covid, vaccine, pre/pro

#CHANGE FILE PATHS

```

#Import files
```{r}
setwd("your_working_directory")
database <- read.csv("PROMOTEDatabase.csv") #Export of the whole database project
database <- database %>% filter(str_detect(id_participant, "PRT"))
database <- database %>% filter(!str_detect(id_participant, "_x"))

longitudinal <- read.csv("PROMOTELongitudinal.csv") #Combined longitudinal projects
longitudinal <- longitudinal %>% filter(!str_detect(id_participant_l, "_x"))

legacy <- read.csv("PQ_SRO_Legacy.csv") #Export of the whole PQ/SRO Legacy project
missingID <- read.csv("MissingIdentifiers.csv") #Missing ID reference file
legacy <- (merge(missingID, legacy, by = 'record_id',  all.y = TRUE))
paste_noNA <- function(x,sep=", ") {
gsub(", " ,sep, toString(x[!is.na(x) & x!="" & x!="NA"] ) ) }
sep=", "
legacy$id_participant<- apply( legacy[ , c(2:3) ] , 1 , paste_noNA , sep=sep)

snq <- read.csv("LegacySNQ.csv") #Export of the whole PNQ Legacy project

sensor <- read.csv("SensorData.csv") #Export of the whole sensor project

covid <- read.csv("covid_long.csv") #Export of the whole original covid project

vaccine <- read.csv("COVID_Vaccine.csv") #Export of the while covid vaccine project

prepro <- read.csv("PrePro2.csv") #Export of the whole pre/probiotic project

pasc <-  read.csv("PASCSurvey.csv") #Export of the whole pasc project

```

#PDDS
##EHR
```{r}
#Database history
pdds_hx <-  database[,c("id_participant", "pdds_number", "pdds_date_1", "pdds_value_1", "pdds_date_2", "pdds_value_2", "pdds_date_3", "pdds_value_3", "pdds_date_4", "pdds_value_4", "pdds_date_5", "pdds_value_5", "pdds_date_6", "pdds_value_6", "pdds_date_7", "pdds_value_7", "pdds_date_8", "pdds_value_8", "pdds_date_9", "pdds_value_9", "pdds_date_10", "pdds_value_10", "pdds_date_11", "pdds_value_11", "pdds_date_12", "pdds_value_12", "pdds_date_13", "pdds_value_13", "pdds_date_14", "pdds_value_14", "pdds_date_15", "pdds_value_15", "pdds_date_16", "pdds_value_16", "pdds_date_17", "pdds_value_17", "pdds_date_18", "pdds_value_18")]
pdds_hx <- pdds_hx %>% filter(!is.na(pdds_number))

pdds_hxScore <- pdds_hx %>% gather(scoreNumber, score, 4,6,8,10,12,14,16,18,20,22,24,26,28,30, 32,34,36,38)
pdds_hxScore <- pdds_hxScore[,c("id_participant", "scoreNumber", "score")]
pdds_hxDate <- pdds_hx %>% gather(dateNumber, date, 3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,35,37)
pdds_hxDate <- pdds_hxDate[,c("id_participant", "dateNumber", "date")]

pdds_hx <- cbind(pdds_hxDate, pdds_hxScore)
pdds_hx <- pdds_hx[,c("id_participant", "date", "score")]
pdds_hx <- pdds_hx %>% filter(!is.na(score))

```

##Blood samples
```{r}
pdds_blood <- database[,c("id_participant", "blood1_date", "blood1_pddsv2")]

pdds_blood <- pdds_blood %>% filter(!is.na(blood1_pddsv2))

names(pdds_blood)[2] <- 'date'
names(pdds_blood)[3] <- 'score'

```

##Stool samples
```{r}
pdds_stool <- database[,c("id_participant", "stool0a_collection_date", "stool0b_collection_date", "collection1a_date", "collection1b_date", "stool_pddsv2_date", "stool_pddsv2")]

paste_noNA <- function(x,sep=", ") {
gsub(", " ,sep, toString(x[!is.na(x) & x!="" & x!="NA"] ) ) }
sep=", "
pdds_stool$date <- apply( pdds_stool[ , c(2:6) ] , 1 , paste_noNA , sep=sep)

pdds_stool <- pdds_stool[,c("id_participant", "date", "stool_pddsv2")]
pdds_stool <- pdds_stool %>% filter(!is.na(stool_pddsv2))
pdds_stool$date <- sub("\\,.*", "", pdds_stool$date)
pdds_stool <- pdds_stool %>% filter(str_detect(date, "20"))

names(pdds_stool)[3] <- 'score'

```

##CSF samples
```{r}
pdds_csf <- database[,c("id_participant", "csf_date", "csf_pddsv2")]

pdds_csf <- pdds_csf %>% filter(!is.na(csf_pddsv2))

names(pdds_csf)[2] <- 'date'
names(pdds_csf)[3] <- 'score'

```

##SRO-Legacy
```{r}
#legacy SRO
pddsLeg <- legacy[,c("id_participant", "promote_self_reported_outcomes_timestamp", "pdds")]
names(pddsLeg)[2] <- 'date'
pddsLeg <- pddsLeg %>% filter(str_detect(date, "20"))
pddsLeg <- pddsLeg %>% filter((!str_detect(pdds, "U")))
names(pddsLeg)[3] <- 'score'


```

##SRO & Update- Long1&2
```{r}
#longitudinal SRO & Update
pddsSROUpd <- longitudinal[,c("id_participant_l", "promote_self_reported_outcomes_timestamp", "pddsv2", "update_questionnaire_timestamp", "pddsv2_4d3bcb")]
pddsSROUpdDate <- pddsSROUpd %>% gather(dateNumber, date, 2,4)
pddsSROUpdScore <- pddsSROUpd %>% gather(scoreNumber, score, 3,5)
pddsSROUpd <- cbind(pddsSROUpdDate, pddsSROUpdScore)
names(pddsSROUpd)[1] <- 'id_participant'

pddsSROUpd <- pddsSROUpd[,c("id_participant", "date", "score")]
pddsSROUpd <- pddsSROUpd %>% filter(!is.na(score))
pddsSROUpd <- pddsSROUpd %>% filter(str_detect(date, "20"))

```

##Outcomes
```{r}
pddsOut <- longitudinal[,c("id_participant_l", "outcome_measures_timestamp", "pddsv2_outcomes")]
names(pddsOut)[2] <- 'date'
pddsOut <- pddsOut %>% filter(str_detect(date, "20"))
names(pddsOut)[1] <- 'id_participant'
names(pddsOut)[3] <- 'score'

```

##Social network legacy
```{r}
#legacy SNQ
pddsSNQ <- snq[,c("redcap_survey_identifier", "social_network_questionnaire_timestamp","social_pdds")]
pddsSNQ <- pddsSNQ %>% filter(str_detect(social_network_questionnaire_timestamp, "20"))
pddsSNQ <- pddsSNQ %>% filter(str_detect(redcap_survey_identifier, "PRT"))
names(pddsSNQ)[1] <- 'id_participant'
names(pddsSNQ)[2] <- 'date'
names(pddsSNQ)[3] <- 'score'


```

##Sensor- monthly
```{r}
#sensor monthly
pddsSensor <- sensor[,c("record_id", "monthlyq_timestamp", "pddsv2")]
pddsSensor <- pddsSensor %>% filter(!str_detect(record_id, "_0"))
pddsSensor <- pddsSensor %>% filter(str_detect(monthlyq_timestamp, "20"))
pddsSensor <- pddsSensor %>% filter(!is.na(pddsv2))
names(pddsSensor)[1] <- 'id_participant'
names(pddsSensor)[2] <- 'date'
names(pddsSensor)[3] <- 'score'

```

##COVID-baseline & quarterly
```{r}
#COVID long, baseline + quarterly
  #Baseline
pddsCovid <- covid[,c("id_participant_l", "covid19_baseline_timestamp", "pddsv2_baseline", "covid19_quarterly_timestamp", "pddsv2_3m")]
pdds_CovidDate <- pddsCovid %>% gather(dateNumber, date, 2,4)
pdds_CovidScore <- pddsCovid %>% gather(scoreNumber, score, 3,5)
pddsCovid <- cbind(pdds_CovidDate, pdds_CovidScore)
names(pddsCovid)[1] <- 'id_participant'
pddsCovid <- pddsCovid[,c("id_participant", "date", "score")]
pddsCovid <- pddsCovid %>% filter(!is.na(score))

```

##COVID vaccine
```{r}
#COVID vaccine
pddsVaccine <- vaccine[,c("id_participant_l", "covid19_vaccine_questions_timestamp", "pddsv2_outcomes")]
pddsVaccine <- pddsVaccine %>% filter(str_detect(covid19_vaccine_questions_timestamp, "20"))
pddsVaccine <- pddsVaccine %>% filter(!is.na(pddsv2_outcomes))
names(pddsVaccine)[1] <- 'id_participant'
names(pddsVaccine)[2] <- 'date'
names(pddsVaccine)[3] <- 'score'
```

##Prebiotic
```{r}
pddsPrePro <- prepro[,c("record_id", "promote_id", "v1_survey_timestamp", "v2_survey_timestamp", "v3_survey_timestamp", "v4_survey_timestamp", "v5_survey_timestamp", "msrsr_and_pdds_questionnaire_timestamp", "pdds")]
pddsPrePro <- pddsPrePro %>%
  group_by(record_id) %>%
  mutate(promote_id = first(promote_id))


paste_noNA <- function(x,sep=", ") {
gsub(", " ,sep, toString(x[!is.na(x) & x!="" & x!="NA"] ) ) }
sep=", "
pddsPrePro$date <- apply( pddsPrePro[ , c(3:8) ] , 1 , paste_noNA , sep=sep)

pddsPrePro <- pddsPrePro %>% filter(!is.na(pdds))
names(pddsPrePro)[2] <- 'id_participant'
names(pddsPrePro)[9] <- 'score'
pddsPrePro <- pddsPrePro[,c("id_participant", "date", "score")]


```

##PASC
```{r}
pasc <- pasc %>% filter(str_detect(id_participant, "PRT"))

pddsPASC <- pasc[,c("id_participant", "covid_date1", "pdds_precovid", "postacute_sequelae_of_sarscov2_timestamp", "pdds_current")]
pddsPASC_date <- pddsPASC %>% gather(dateType, date, 2,4)
pddsPASC_date <- pddsPASC_date[,c(1,4,5)]

pddsPASC_score <- pddsPASC %>% gather(scoreType, score, 3,5)
pddsPASC_score <- pddsPASC_score[,c(1,4,5)]

pddsPASC2 <- cbind(pddsPASC_date, pddsPASC_score)
pddsPASC2 <- pddsPASC2[,c(1,3,6)]

```


##Combine all PDDS
```{r}
pdds_all <- rbind(pdds_hx, pdds_blood, pdds_stool, pdds_csf, pddsLeg, pddsSROUpd, pddsOut, pddsSNQ, pddsSensor, pddsCovid, pddsVaccine, pddsPASC2, pddsPrePro)
  #remove pddsCovid for AR-PDDS score compilation
pdds_all <- pdds_all %>% filter(str_detect(date, "20"))
pdds_all <- pdds_all %>% filter(!is.na(score))
pdds_all <- pdds_all %>% filter(score!=9)
pdds_all <- unique(pdds_all)

pdds_all$score <- as.numeric(pdds_all$score)
pdds_all$date <- as.Date(pdds_all$date)
  #filter for scores through 2022 for the data desc paper
pdds_all <- pdds_all %>% filter(str_detect(id_participant, "PRT"))
  #filter out bad dates- greater than today
pdds_all <- pdds_all %>% filter(!date>Sys.Date())


test <- data.frame(table(pdds_all$id_participant))
test <- test %>% filter(Freq>0)
  #2105 participants with scores
mean(test$Freq) #7.4 scores per participant
sd(test$Freq)

```

#PDDS Visualization
##Histogram of all pdds scores
```{r}
#Export list of all pdds scores
#write.csv(pdds_all, "/Users/LibbyWalker/Documents/Job Stuff/UPitt/R/Data Desc Queries/PDDS_Query/pdds_all.csv")

mean(pdds_all$score) #mean score of 2.04
sd(pdds_all$score) #sd of 2.17

ggplot(pdds_all, aes(score)) +
    geom_histogram(colour="#4d3230", fill="#c2a2a0", bins = 9) +
    stat_bin(aes(y=..count.., label=ifelse(..count..==0,"",..count..)), geom="text", vjust=-.5) + 
    theme_minimal() +
    geom_vline(aes(xintercept=mean(score)), color="#304b4d", linetype="dashed", size=.5) +
    scale_x_continuous(breaks = seq(0, 9, by=1)) +
    scale_y_continuous(breaks = seq(0, 6000, by=500)) +
    labs(title = "All PDDS Scores") + theme(plot.title = element_text(hjust = 0.5)) + xlab("PDDS Score") + ylab("# of Scores")



```

##Histogram of most recent pdds scores
```{r}
most_recent <- pdds_all
most_recent = aggregate(pdds_all$date,by=list(most_recent$id_participant),max)
names(most_recent)[1] <- 'id_participant'
names(most_recent)[2] <- 'date'
  #merge list of most recent pdds dates with the scores
most_recent2 = merge(most_recent, pdds_all, by.x=c("id_participant", "date"), by.y=c("id_participant", "date"))
most_recent2 <- unique(most_recent2)

  #find duplicate PRTs- multiple scores on the same date
test <- data.frame(table(most_recent2$id_participant))
test <- test %>% filter(Freq>0)
  #remove lower score, only 1 per participant, total should be 1921
most_recent3 <- most_recent2 %>% filter( (!((score==0) & (str_detect(id_participant, "PRT170109")))))
most_recent3 <- most_recent3 %>% filter( (!((score==1) & (str_detect(id_participant, "PRT170218")))))
most_recent3 <- most_recent3 %>% filter( (!((score==0) & (str_detect(id_participant, "PRT180558")))))
most_recent3 <- most_recent3 %>% filter( (!((score==3) & (str_detect(id_participant, "PRT180814")))))
most_recent3 <- most_recent3 %>% filter( (!((score==0) & (str_detect(id_participant, "PRT191184")))))
most_recent3 <- most_recent3 %>% filter( (!((score==0) & (str_detect(id_participant, "PRT191202")))))
most_recent3 <- most_recent3 %>% filter( (!((score==0) & (str_detect(id_participant, "PRT191296")))))
most_recent3 <- most_recent3 %>% filter( (!((score==1) & (str_detect(id_participant, "PRT191315")))))
most_recent3 <- most_recent3 %>% filter( (!((score==0) & (str_detect(id_participant, "PRT191580")))))
most_recent3 <- most_recent3 %>% filter( (!((score==2) & (str_detect(id_participant, "PRT201657")))))
most_recent3 <- most_recent3 %>% filter( (!((score==1) & (str_detect(id_participant, "PRT212128")))))
most_recent3 <- most_recent3 %>% filter( (!((score==1) & (str_detect(id_participant, "PRT212258")))))
most_recent3 <- most_recent3 %>% filter( (!((score==3) & (str_detect(id_participant, "PRT222673")))))
test <- data.frame(table(most_recent3$id_participant))
test <- test %>% filter(Freq>0)

mean(most_recent3$score) #mean score of 2.08
sd(most_recent3$score)

ggplot(most_recent3, aes(score)) +
    geom_histogram(colour="#4d3230", fill="#c2a2a0", bins = 9) +
    stat_bin(aes(y=..count.., label=ifelse(..count..==0,"",..count..)), geom="text", vjust=-.5) + 
    theme_minimal() +
    geom_vline(aes(xintercept=mean(score)), color="#304b4d", linetype="dashed", size=.5) +
    scale_x_continuous(breaks = seq(0, 9, by=1)) +
    scale_y_continuous(breaks = seq(0, 6000, by=500)) +
    labs(title = "Most Recent PDDS Scores") + theme(plot.title = element_text(hjust = 0.5)) + xlab("PDDS Score") + ylab("# of Scores")

#Export list of recent pdds scores
#write.csv(most_recent3, "/Users/LibbyWalker/Documents/Job Stuff/UPitt/R/Data Desc Queries/PDDS_Query/pdds_recent.csv")


```

##Histogram of earliest pdds scores
```{r}
earliest <- pdds_all
earliest = aggregate(pdds_all$date,by=list(earliest$id_participant),min)
names(earliest)[1] <- 'id_participant'
names(earliest)[2] <- 'date'
  #merge list of earliest pdds dates with the scores
earliest2 = merge(earliest, pdds_all, by.x=c("id_participant", "date"), by.y=c("id_participant", "date"))
earliest2 <- unique(earliest2)

  #find duplicate PRTs- multiple scores on the same date
test <- data.frame(table(earliest2$id_participant))
test <- test %>% filter(Freq>0)
  #remove lower score, only 1 per participant, total should be 1921
earliest3 <- earliest2 %>% filter( (!((score==0) & (str_detect(id_participant, "PRT170107")))))
earliest3 <- earliest3 %>% filter( (!((score==1) & (str_detect(id_participant, "PRT170387")))))
earliest3 <- earliest3 %>% filter( (!((score==3) & (str_detect(id_participant, "PRT180814")))))
earliest3 <- earliest3 %>% filter( (!((score==1) & (str_detect(id_participant, "PRT212128")))))
earliest3 <- earliest3 %>% filter( (!((score==3) & (str_detect(id_participant, "PRT222673")))))
test <- data.frame(table(earliest3$id_participant))
test <- test %>% filter(Freq>0)

mean(earliest3$score) #mean score of 1.81
sd(earliest3$score)

ggplot(earliest3, aes(score)) +
    geom_histogram(colour="#ad908e", fill="#c2a2a0", bins = 9) +
    theme_linedraw() +
    geom_vline(aes(xintercept=mean(score)), color="#E74C3C", linetype="dashed", size=.5) +
    scale_x_continuous(breaks = seq(0, 9, by=1)) +
    scale_y_continuous(breaks = seq(0, 1000, by=200)) +
    theme(plot.title = element_text(hjust = 0.5)) + xlab("PDDS Score") + ylab("Number of Participants") + theme(axis.title.x = element_text(margin = margin(t = 15)), axis.title.y = element_text(margin = margin(r=15)))

#Export list of earliest pdds scores
#write.csv(earliest3, "/Users/LibbyWalker/Documents/Job Stuff/UPitt/R/Data Desc Queries/PDDS_Query/pdds_earliest.csv")


```

##Minimum two pdds scores
```{r}
#Finds list of participants with at least 2 MSRSR scores on separate days
#Then displays histogram of the days between the two scores

multiple <- data.frame(table(pdds_all$id_participant))
multiple <- multiple %>% filter(Freq>1)
names(multiple)[1] <- 'id_participant'
  #merge list of >1 score with the date and actual scores
multiple2 <- merge(multiple, pdds_all, by = 'id_participant', all.y = TRUE)
multiple2 <- multiple2 %>% filter(!is.na(Freq))
multiple2 <- multiple2[order(multiple2$id_participant, multiple2$date), ]

  #calculate the difference in dates by days, absolute value
multiple3 <- multiple2 %>%
  group_by(id_participant) %>%
  mutate(diff = lag(date, default = first(date), order_by = Freq) - date)
multiple3$diff <- abs(multiple3$diff)
  #calculate the mean of the difference in dates- any mean that ==0 has only scores from the same day
test2 <- multiple3 %>%
  group_by(id_participant) %>%
  summarise_at(vars(diff), list(mean_days_diff = mean))
  #remove people w/ mean_days_diff==0
test2 <- test2 %>% filter(mean_days_diff!=0)

sum(table(unique(test2$id_participant)))
  #1701 participants w/ 2+ scores on 2+ days
mean(test2$mean_days_diff) #133.35 days between scores per person

test2$mean_days_diff <- as.numeric(as.character(test2$mean_days_diff))
ggplot(test2, aes(mean_days_diff)) +
    geom_histogram(colour="#ad908e", fill="#c2a2a0") +
    theme_minimal() +
    geom_vline(aes(xintercept=mean(mean_days_diff)), color="#304b4d", linetype="dashed", size=.5) +
    scale_y_continuous(breaks = seq(0, 300, by=50)) +
    scale_x_continuous(breaks = seq(0, 750, by=60)) +
    labs(title = "Mean Duration Between Scores per Participant") + theme(plot.title = element_text(hjust = 0.5)) + xlab("Duration (days)") + ylab("# of Participants")

```

#PDDS Median
```{r}
#Filtering values
pdds_final <- pdds_all
pdds_final$year <- year(pdds_final$date)
pdds_final_summary <- pdds_final %>% 
  mutate_at(vars(-id_participant), as.numeric) %>%
  group_by(id_participant, year) %>%
  dplyr::summarise(pdds_median = floor(median(score)),
            pdds_num_observations = n()) %>% ungroup()
    #groups table by participant ID and year, then finds the median value.  So for one person, in 2017, the median is found.

pdds_new <- merge(pdds_final, pdds_final_summary, by = c("id_participant", "year"))
pdds_new$difference <- pdds_new$score - pdds_new$pdds_median
    #calculates the difference between each individual score and the median for that year
#Removing 1.5 x (IQR/2) above or below the median
pdds_new <- pdds_new %>% filter(!(pdds_new$difference > 3 | pdds_new$difference < -3))
pdds_new <- pdds_new[,c(1,2,3,4)]
    #removes values that are more than 3 points off from the median for that participant for that year

pdds_new_summary <- pdds_new %>% 
  mutate_at(vars(-id_participant), as.numeric) %>%
  group_by(id_participant, year) %>%
  dplyr::summarise(pdds_median = floor(median(score)),
            pdds_num_observations = n()) %>% ungroup()
  #recalculates the median having removed outliers (values way greater than the median for that year)

pdds_new_summary <- pdds_new_summary %>%
  group_by(id_participant) %>%
  mutate(diff = lag(pdds_median, default = first(pdds_median), order_by = year) - pdds_median)
pdds_new_summary <- as.data.frame(pdds_new_summary)

#export list of all median scores per year
pdds_all_median <- pdds_new_summary
write.csv(pdds_all_median, "pdds_all_median.csv")


```

#EDSS
```{r}
edss <- database[,c("id_participant", "edss_date", "edss_score")]
edss <- edss %>% filter(!is.na(edss_score))
write.csv(edss, "edss_all.csv")

hist(edss$edss_score)

```



